---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/Shahil Sunder/Desktop/Reshay_Project/coursera/ProgrammingAssignment2/RepData_PeerAssessment 1/RepData_PeerAssessment1")
library(tidyr)
library(dplyr)
library(ggplot2)
library(xtable)
```
## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:
- steps: Number of steps taking in a 5-minute interval (missing values are coded as \color{red}{\verb|NA|}NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Requirements:

1. Code for reading in the dataset and/or processing the data
2. Histogram of the total number of steps taken each day
3. Mean and median number of steps taken each day
4. Time series plot of the average number of steps taken
5. The 5-minute interval that, on average, contains the maximum number of steps
6. Code to describe and show a strategy for imputing missing data
7. Histogram of the total number of steps taken each day after missing values are imputed
8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report


## Loading and preprocessing the data

The csv file containing the activity data is loaded into a data frame called actData using the code below.The structure of the data, dimensions as well as the first and last 10 rows are viewed

```{r activity data, echo=TRUE}
        actData <- data.frame(read.csv("activity.csv", header = TRUE))
        str(actData)
        dim(actData)
        head(actData, 5)
        tail(actData, 5)
        missingData <- actData[is.na(actData),]
        dim(missingData)
```

The data loaded is further processed to remove the missing values. The data is store in a data frame named actDataComplete.

```{r activity data complete cases, echo=TRUE}
        actDataComplete <- data.frame(actData[complete.cases(actData),])
        str(actDataComplete)
        dim(actDataComplete)
        head(actDataComplete, 5)
        tail(actDataComplete, 5)
```

## What is mean total number of steps taken per day?

The total number of steps take per day is determined using the tapply method, summing the steps of the "actDataComplete" data, as a factor of the date. The complete cases, are only considered.

```{r total steps per day, echo=TRUE}
        totalSteps <- group_by(actDataComplete, date)
        totalSteps <- summarise(totalSteps, dailySteps = sum(steps))
        head(totalSteps, 5)
        mean(totalSteps$dailySteps)
```

The histogram illustrating the "Total number of steps taken each day" is generated in the base plotting system.

```{r histogram: total steps per day, echo=TRUE, fig.height=7, fig.width= 7, fig.align= 'center'}

        hist(totalSteps$dailySteps, main = "Total number of steps taken each day", xlab = "Total Steps per day")
```

The mean and median are found in the summary below:

```{r xtable, results='asis'}
      sumTbl <- data.frame(Mean=mean(totalSteps$dailySteps), Median = median(totalSteps$dailySteps), stringsAsFactors=FALSE) 
      sumTbl <- xtable(sumTbl, digits = c(0,0,0))
      print(sumTbl, type ="html")
      
  
```
## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
